### 2025-09-18 paper/sec/introduction.tex
Merge some related work.

Making sense of data-driven claims is hard, even if the evidence base (data sets and related analyses) is
open. We see this in peer review, misinformation, retracted papers, \rpnote{supporting citations}.
\rpnote{Small example here?}

Recent work on data provenance and data visualisation~\citep{psallidas18smoke,perera22,bond25} proposed more
transparent charts and figures by integrating interactive ``provenance queries'' directly into visual outputs.
These interactive queries allow readers to interact with visual elements to see the data relevant to those
elements. The key feature of these approaches is that the interactions are provided automatically, via
\emph{transparent programming languages} (either query languages or general-purpose languages) which track
dependencies as data flows through a computation.

\begin{itemize}
    \item Making the text transparent, with the possibility to highlight the data related to a specific sentence.
\end{itemize}

\subsection{Transparent Text via Interactive Provenance Queries}

In most articles, however, charts and figures usually only play a supporting role. The main claims are usually
presented in natural language, language which also makes quantitative statements and refers to ``behind the
scenes'' data analyses. In principle transparent programming languages can also be used to create natural
language which is bidirectly linked to data, by computing key fragments of natural language from data. The
following example, adapted from \cite{moosavi21}, shows how the language presented in this paper can be used
to achieve this:

\begin{figure}[h]
    \centering
    \includegraphics[width=\linewidth]{fig/scigen-1805.02474v1-10-with-pointer.png}
    \caption{\emph{Transparent text} links fragments of natural language to underlying data, via queries}\label{fig:scigen-example-website}
\end{figure}

>>>>

Making sense of and verifying data-driven claims is hard, even when the evidence base (data sets and related
analyses) is open. We see this in peer review, misinformation, retracted papers, \rpnote{supporting
citations}. The ability of LLMs to assist with these challenges has received considerable attention recently,
for example in quantity extraction~\citep{bolucu23}, scientific fact-checking~\citep{abu-ahmad25}, and
scientific figure interpretation~\citep{roberts24}.

Recent work on data provenance and data visualisation~\citep{psallidas18smoke,perera22,bond25} attempts to
address this issue through software infrastructure for ``transparent'' charts and figures. These are visual
outputs that provide interactive \emph{provenance queries}, user interactions (e.g. mousing over a visual
element) that allow readers to explore how visual features relate to data. In these approaches the
interactions are provided automatically via an underlying programming language (either a query language or
general-purpose language) which tracks dependencies as data flows through a computation.

In this paper, we extend interactive provenance queries to the natural language that accompanies the charts
and figures. We believe this is an important step because in most articles, charts and figures usually only
play a supporting role. The main claims are usually presented in natural language, language which also makes
quantitative statements and contains substantial implicit analytic content. In principle transparent
programming languages can also be used to create natural language which is bidirectly linked to data, by
computing key fragments of natural language from data. The following example, adapted from \cite{moosavi21},
shows how the language presented in this paper can be used to achieve this:

\begin{figure}%[h]
    \centering
    \includegraphics[width=\linewidth]{fig/scigen-1805.02474v1-10-with-pointer.png}
    \caption{\emph{Transparent text} links fragments of natural language to underlying data, via queries}\label{fig:scigen-example-website}
\end{figure}
